Abstractions and exemplars: the measure noun phrase alternation in German

### 2018-01-03 Latex/sections/5conclusions.tex
Final fixes and anonymisation.

The empirical studies in Sections~\ref{sec:corpusstudies} and~\ref{sec:experimental} confirm the theoretical model proposed in Section~\ref{sec:germanmeasurenps}, which predicted prototype effects and exemplar effects to jointly determine which of the two alternating constructions is chosen.
More specifically, the pre-study and the main study provided evidence that the NAC constructions favour more strongly grammaticalised measure nouns, modification by cardinals, and styles where the genitive is underrepresented and the text is less coherently and elaborately written, as captured in the \textit{Badness} variable.%
\footnote{Although, as pointed out in Section~\ref{sec:mainstudyinterpretation}, the \textit{Badness} effect is weak and not detected in the pre-study, probably as a consequence of its weakness.}
The overall convergence between the pre-study and the main study and also the expected result for \textit{Measurecase} given previous research \citep{Zimmer2015} -- although it was only a control variable here -- strengthen the validity of this study even without the experimental validation.

While the results do not \textit{prove} -- due to the intrinsic equivalence of prototype and exemplar models with respect to the output they produce -- that the model proposed in Section~\ref{sec:germanmeasurenps} is congruent with any speaker's mental representation, the two attraction features (\textit{Kindattraction} and \textit{Measureattraction}) are virtually impossible to conceive of as prototype effects.
As discussed in Section~\ref{sec:cogocl}, it is unnecessary to follow an extreme route, be it radical prototype theory or radical exemplar theory given recent developments in cognitive science and (less prominently) cognitive linguistics.
While it is surely an intentionally overstated comment, \citet[15]{Kapatsinski2014} suggests that ``[i]n the extreme, some speakers’ heads could host exemplar models, and some could contain fairly abstract grammars, and the produced output would be essentially identical''.
If this is (to a probably less extreme degree) the case, pooled data from corpora might only provide (still highly informative) data averaged over levels of abstraction.
Clearly, more research is required in this direction.

@@

\begin{quote}
  There are now a number of published multivariate models that use data, extracted from corpora [\ldots] to predict the choice for one morpheme, lexeme or construction over another.
  However, [\ldots] only a small number of these corpus-based studies have been cross-validated [\ldots].
  Of these cross- validated studies, few have directly evaluated the prediction accuracy of a complex, multivariate corpus-based model on humans using authentic corpus sentences [\ldots].
\end{quote}

The question now arises whether convergence was reached in the present case.
The answer is a clear yes.
The predictions made by the corpus-based model were a significant factor, both in the more explicit paradigm (forced choice) and the implicit paradigm (self-paced reading).
This shows that the model indeed predicts the variant which language users expect.
The overall effect as measured in the $R^2$ was weak (roughly 0.2) in both cases, however.
While part of this could be traced back to a suboptimally chosen stimulus, we really need to consider what kind of convergence we expect to see between corpus data and experiments.
The main corpus study had $R^2_m=0.409$ and $R^2_c=0.495$, which is good but not anywhere near a perfect fit.
With the added inter-speaker variability brought into play by the experimental setup (compared to the averaging across thousands of speakers in the corpus study), a perfect fit cannot be expected.
Like many phenomena in German which are often called \textit{Zweifelsfälle} (`cases of doubt') in the traditional literature \citep{Duden09,Klein2009} the MNP alternation is a case where speakers very often have no clear intuition and a lot of free variation seems to be involved.
Rarely do speakers feel that one alternative is clearly odd or bad.
Additionally, cases of doubt involving the genitive are often a matter of fierce public debate, and especially the forced choice paradigm does not effectively prevent participants from making normative judgements.
This might even account for the slightly better fit in the self-paced reading experiment, where normative considerations are suppressed.
Thus, the present study shows that what counts as convergence between corpus and experimental results should be gauged considering the nature of the phenomenon at hand, the source of the corpus data, and the experimental paradigm.
Clearly, more case studies using diverse and different corpora are needed, and it should become standard practice to cross-validate them using experimental methods.
Given that a single failure to achieve convergence does not provide any evidence for divergence, many more studies need to be published to the point that meta-analysis becomes possible.%
\footnote{This is even more vital considering the variation in results from experimental work.
See, for example, the impressive list of different reading time results from ten papers on Chinese relative clause processing in \citet[8]{Vasishth2015}.}

On a larger methodological scale, this paper also makes a number of contributions.
The statistical model was a true multilevel model (see Section~\ref{sec:itemandexemplareffects} and Section~\ref{sec:mainstudystatisticalmodel}), demonstrating how multilevel modelling helps to specify complex hierarchies of abstract features, item-specific tendencies, and exemplar effects at the level of observations (sentences; first level) and lemmas (second level).
While \cite{Gries2015} still calls mixed models ``underused'' in corpus linguistics, multilevel models are at least equally underused tools.
At the same time, the effects of overparametrisation of mixed models with varying slopes as criticised by \cite{BatesEa2015a} were demonstrated.
Furthermore, fitting an additive model to the reading time data did not improve the fit as there were no non-linearities in the data.
While \cite{DivjakEa2016} also use attested sentences, they find that an additive model helped to deal with non-linearities.
This is clearly another area where only more studies can lead to clarification.
Finally, much like \cite{Dabrowska2014} found that speakers' knowledge of collocations was not matched by a set of standard measures of collocation strength extracted from corpora, a simple quotient based on raw frequencies for the attraction strength performed much better in the present study compared to collexeme strength (see Section~\ref{sec:corpusstudies}).
While this does not allow the conclusion that collexeme strength has no cognitive reality, it might indicate that for measuring attraction effects exerted by non-alternating constructions on alternating constructions, other measures might be more appropriate.

>>>>

The empirical studies in Sections~\ref{sec:corpusstudies} and~\ref{sec:experimental} confirm the theoretical model proposed in Section~\ref{sec:germanmeasurenps}, which predicted prototype effects and exemplar effects to jointly determine which of the two alternating constructions is chosen.
More specifically, the pre-study and the main study provided evidence that the NAC constructions favour more strongly grammaticalised measure nouns, modification by cardinals, and styles where the genitive is underrepresented and the text is less coherently and elaborately written, as captured in the \textit{Badness} variable.%
\footnote{Although, as pointed out in Section~\ref{sec:mainstudyinterpretation}, the \textit{Badness} effect is weak and not detected in the pre-study, probably as a consequence of its weakness.}
The overall convergence between the pre-study and the main study and also the expected result for \textit{Measurecase} in line with previous research \citep{Zimmer2015} -- although it was only a control variable here -- strengthen the validity of this study even without the experimental validation.

While the results do not \textit{prove} -- due to the intrinsic equivalence of prototype and exemplar models with respect to the output they produce -- that the model proposed in Section~\ref{sec:germanmeasurenps} is congruent with any speaker's mental representation, the two attraction features (\textit{Kindattraction} and \textit{Measureattraction}) are virtually impossible to conceive of as prototype effects.
As discussed in Section~\ref{sec:cogocl}, it is unnecessary to follow an extreme route, be it radical prototype theory or radical exemplar theory given recent developments in cognitive science and (less prominently) cognitive linguistics.
While it is surely an intentionally overstated comment, \citet[15]{Kapatsinski2014} even suggests that ``[i]n the extreme, some speakers’ heads could host exemplar models, and some could contain fairly abstract grammars, and the produced output would be essentially identical''.
If this is (to a probably less extreme degree) the case, corpora would only provide (still informative) pooled data averaged over varying levels of abstraction.
Clearly, more research is required in this direction.

@@

\begin{quote}
  There are now a number of published multivariate models that use data[,] extracted from corpora [\ldots] to predict the choice for one morpheme, lexeme or construction over another.
  However, [\ldots] only a small number of these corpus-based studies have been cross-validated [\ldots].
  Of these cross- validated studies, few have directly evaluated the prediction accuracy of a complex, multivariate corpus-based model on humans using authentic corpus sentences [\ldots].
\end{quote}

The question now arises whether convergence was reached in the present case or not.
The answer is a clear yes.
The predictions made by the corpus-based model were a significant factor, both in the more explicit paradigm (forced choice) and the implicit paradigm (self-paced reading).
This shows that the model indeed predicts the variant which language users expect.
The overall effect as measured in the $R^2$ was weak (roughly 0.2) in both cases, however.
While part of this could be traced back to one suboptimally chosen stimulus, we really need to consider what kind of convergence we expect to see between corpus data and experiments.
The main corpus study had $R^2_m=0.409$ and $R^2_c=0.495$, which is good but not anywhere near a perfect fit.
With the added inter-speaker variability brought into play by the experimental setup (compared to the averaging across thousands of speakers in the corpus study), a perfect fit cannot be expected.
Like many phenomena in German which are often called \textit{Zweifelsfälle} (`cases of doubt') in the traditional literature \citep{Duden09,Klein2009} the MNP alternation is a case where speakers very often have no clear intuition and a lot of free variation seems to be involved.
Rarely do speakers feel that one alternative is clearly odd or bad.
Additionally, cases of doubt involving the genitive are often a matter of fierce normative public debate, and especially the forced choice paradigm does not effectively prevent participants from making normative judgements.
This might even account for the slightly better fit in the self-paced reading experiment, where normative considerations are suppressed.
Thus, the present study shows that what counts as convergence between corpus and experimental results should be gauged considering the nature of the phenomenon at hand, the source of the corpus data, and the experimental paradigm.
Clearly, more case studies using diverse and different corpora are needed, and it should become standard practice to cross-validate them using experimental methods.
Given that a single failure to achieve convergence does not provide any evidence for divergence, many more studies need to be published to the point that meta-analysis becomes possible.%
\footnote{This is even more vital considering the variation in results from experimental work.
See, for example, the impressive list of different reading time results from ten papers on Chinese relative clause processing in \citet[8]{Vasishth2015}.}

On a larger methodological scale, this paper also makes a number of contributions.
The statistical model was a true multilevel model (see Section~\ref{sec:itemandexemplareffects} and Section~\ref{sec:mainstudystatisticalmodel}), demonstrating how multilevel modelling helps to specify complex hierarchies of abstract features, item-specific tendencies, and exemplar effects at the level of observations (sentences; first level) and lemmas (second level).
While \cite{Gries2015} still calls mixed models ``underused'' in corpus linguistics, multilevel models are consequently at least equally underused tools.
At the same time, the effects of overparametrisation of mixed models with varying slopes as criticised by \cite{BatesEa2015a} were demonstrated.
Furthermore, fitting an additive model to the reading time data did not improve the fit as there were no non-linearities in the data.
While \cite{DivjakEa2016} also use attested sentences, they find that an additive model helped to deal with non-linearities.
This is clearly another area where only more studies can lead to clarification.
Finally, much like \cite{Dabrowska2014} found that speakers' knowledge of collocations was not matched by a set of standard measures of collocation strength extracted from corpora, a simple quotient based on raw frequencies for the attraction strength performed much better in the present study compared to collexeme strength (see Section~\ref{sec:corpusstudies}).
While this does not allow the conclusion that collexeme strength has no cognitive reality, it might indicate that for measuring attraction effects exerted by non-alternating constructions on alternating constructions, other measures might be more appropriate.
